Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-02 paper.tex

The \texttt{DemographyDebugger} can output detailed information for debugging
demographic models, but also provides
methods to compute a suite of numerical predictions about the location of
lineages over time, for example the coalescence rates for two or more lineages
drawn from populations at specified times in the past,
which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
or IICR of \citet{chikhi2018inverse}.
Many popular approaches in population genetics use the distribution of
coalescence rates between pairs of lineages in one or more populations to infer
effective population sizes over
time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
split times and subsequent migration rates between
populations~\citep{wang2020tracking}.
The \texttt{DemographyDebugger} provides a valuable
ground-truth when evaluating such inference methods, as illustrated by
\cite{adrion2020community}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} 
representation of the coalescent with recombination in terms of a graph. 
This formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation, 
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows.
We may consider the big ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral material and
consequently wait an exponentially large time until all sampled lineages
coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
which we only track genetic ancestors, and wait until all marginal trees
have fully coalesced (like Hudson's algorithm).

>>>>

The \texttt{DemographyDebugger} can output detailed information for debugging
demographic models, but also provides
methods to compute a suite of numerical predictions about the location of
lineages over time, for example the coalescence rates for two or more lineages
drawn from populations at specified times in the past,
which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
or IICR of \citet{chikhi2018iicr}.
Many popular approaches in population genetics use the distribution of
coalescence rates between pairs of lineages in one or more populations to infer
effective population sizes over
time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
split times and subsequent migration rates between
populations~\citep{wang2020tracking}.
The \texttt{DemographyDebugger} provides a valuable
ground-truth when evaluating such inference methods, as illustrated by
\cite{adrion2020community}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination in terms of a graph.
This formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows.
We may consider the big ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral material and
consequently wait an exponentially large time until all sampled lineages
coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
which we only track genetic ancestors, and wait until all marginal trees
have fully coalesced (like Hudson's algorithm).
